---
title: "Getting Started in R and RStudio"
subtitle: "Parts 1 & 2 (4/30/26, 5/11/26)"
pagetitle: "Code for Getting Started in R and RStudio"
institute: "OHSU OCTRI-BERD"
author: "Jessica Minnier, Meike Niederhausen"
date: "today"
format: 
  html:
    link-external-newwindow: true
    toc: true
    toc_float:
      collapsed: false
    number-sections: true
    number-depth: 4
    theme: lumen
    self-contained: true
    html-math-method: mathjax
execute:
  echo: true
editor: visual
editor_options: 
  chunk_output_type: inline
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(writexl)
library(rstatix)
library(janitor)
library(naniar)
library(gt)
library(gtsummary)
library(readxl)  # part of tidyverse, but not the tidyverse core
```

# Load files for today's workshop

1.  Open Google Doc <https://bit.ly/berd_r_intro_2024_doc>

    - Use this to post questions!
    - This has all the links in it

2.  Download zipped folder of files using link <https://bit.ly/berd_intro_2024_files>

    - **Do you have a PC? Need to right click .zip file and select EXTRACT ALL**

    - Do you have a Mac? Just double click .zip file and it will open

    - These files are hosted on github: <https://github.com/OHSU-OCTRI-BERD/R_Intro_2024_09>

3.  Slides are here: <https://bit.ly/berd_intro_2024_slides> or you can open the slides file (html) in a web browser: `slides_R_berd_2024.html`

4.  [Open RStudio by double-clicking on the file]{style="color:darkorange"} `berd_r_intro_2024_09.Rproj`

## Overview

### Part 1: 09/19/2024

- Intro to R & RStudio
- Quarto documents & code chunks
- R packages
- Data: loading, viewing
- Summarizing data
  - Quantitative
  - Categorical
- Data visualizations
  - Quantitative: histograms, boxplots, scatterplots
  - Categorical: barplots
- Missing data summary

### Part 2: 10/03/2024

- Data wrangling
  - Sorting data
  - Subsetting data (rows, columns)
  - Creating new variables in a dataset
  - Factor level variables
- Some more data viz (stratifying figures)
- Replacing missing values with another value
- Resources

# R Packages: install & load

## Install packages used in this workshop

- Run the code below to install packages:

```{r}
#| eval: FALSE
install.packages("tidyverse")  # this takes a while to install
install.packages("writexl")
install.packages("rstatix") 
install.packages("janitor")
install.packages("naniar")
install.packages("gt")
install.packages("gtsummary")
```

- It is **VERY IMPORTANT** that you install these packages in order to run the code for the workshop!

## Load packages with `library()` command

- Use the `library()` command to load each required package.
- Packages need to be loaded *every* time you open Rstudio.
  - **The code to load packages MUST be in your qmd file.**
  - *Tip: at the top of your qmd file, create a chunk that loads all of the R packages you want to use in that file.*

```{r}
library(janitor)    # run this every time you open Rstudio
```

- Check to see whether a package has been loaded:
  - Go to Packages tab
  - If box to left of package name is checked off, then it's loaded

## Load packages used in this workshop

- Run the code below in your R session

```{r}
library(tidyverse)
library(writexl)
library(rstatix)
library(janitor)
library(naniar)
library(gt)
library(gtsummary)
library(readxl)  # part of tidyverse, but not the tidyverse core
```

- Note that the package names do not need quotes around them when using the `library()` function, but they are required when using the `install.packages()` function.

# Data basics in R

## Dataset example (toy data)

- Use Excel to open the dataset `toy_data.xlsx` that is located in the data folder.

## Load data from an .xlsx (Excel) file: **point & click option**

- In RStudio's Files window, open the `data` folder and
  - then click on the dataset `toy_data.xlsx`.
- Select the `Import Dataset...` option,
  - Import Options: Type in `NA` in the "NA:" box, and
  - then click on the `Import` button on the bottom right of the pop-up window.
- **Copy the code from the console used to load the file into your script** (omitting the + sign):

```{r}
toy_data <- read_excel("data/toy_data.xlsx", na = "NA")
```

## Load data from an .xlsx (Excel) file: **code option**

- The point and click option used the `read_excel` command from the `readxl` package to load the xlsx file.
- When loading the dataset, we need to specify where the xlsx file is located, which in this case is in the folder called `data`.
- No matter which option we use, the [**code to load the data MUST be in your qmd**]{style="color:darkorange"}.

```{r}
# library(readxl) # commented out since loaded above already
toy_data <- read_excel("data/toy_data.xlsx", na = "NA")
```

- This code assigns the output of the `read_excel()` function (data) into an object named `toy_data`
- The [`<-` is the assignment operator]{style="color:green"}
- After loading, the name of the data frame object `toy_data` will be in your Environment tab

## View the `toy_data` dataset

- Run the `View` command below within RStudio.

```{r}
#| eval: FALSE
View(toy_data)
```

- A new tab in the scripting window will appear with the `toy_data` dataset.
- Note that the code chunk in the qmd file specifies `#| eval: FALSE` so that this code is not run while rendering the qmd file.

## What are the variable "types"?

| R variable type  | Description                                          |
|------------------|------------------------------------------------------|
| `dbl`: double    | numbers; also `num` or `int` for numbers or integers |
| `chr`: character | text, "strings"                                      |
| `fct`: factor    | categorical variables stored with levels (groups)    |
| `lgl`: logical   | boolean (TRUE, FALSE)                                |

 

```{r}
glimpse(toy_data)
```

## Getting to know the dataset

### How many rows and columns?

```{r}
# dim is for dimensions
# Output: # of rows, # of columns
dim(toy_data)   

nrow(toy_data)

ncol(toy_data)
```

### What are the variable names?

```{r}
names(toy_data)
```

# The Youth Risk Behavior Surveillance System (YRBSS)

Data from the CDC's [Youth Risk Behavior Surveillance System (YRBSS)](https://www.cdc.gov/healthyyouth/data/yrbs/index.htm)

- complex survey data
- national school-based survey
  - conducted by CDC and state, territorial, and local education and health agencies and tribal governments
- monitors six categories of health-related behaviors
  - that contribute to the leading causes of death and disability among youth and adults
  - including alcohol & drug use, unhealthy & dangerous behaviors, sexuality, and physical activity
  - see [Questionnaires](https://www.cdc.gov/healthyyouth/data/yrbs/questionnaires.htm)
- the data in `yrbss_demo.xlsx` are a subset of data in the R package [`yrbss`](https://github.com/hadley/yrbss), which includes YRBSS from 1991-2013

# Practice time! (part 1)

Use R code to answer the following questions.

1.  Open the Excel file `yrbss_demo.xlsx` in the *data* *folder* to familiarize yourself with it.
2.  Load the dataset `yrbss_demo.xlsx`. Make sure the name of the loaded data is `yrbss_data`.

```{r}

```

3.  What are the column names?

```{r}

```

4.  How many rows and columns are in `yrbss_data`?

```{r}

```

5.  Which variables are numeric (quantitative)?

```{r}

```

6.  Which variables are character?

```{r}

```

# Load the YRBSS data

- Open the Excel file `yrbss_demo.xlsx` in the *data* *folder* to familiarize yourself with it.
- This is a larger version of the toy data above, but it is still a subset of the full data.

```{r}
# library(readxl)
# note we are saving the data from the Excel file in our environment 
# with the object name yrbss_data
yrbss_data <- read_excel("data/yrbss_demo.xlsx")
```

First look at the data:

```{r}
glimpse(yrbss_data)
```

- After loading the dataset, you will see `yrbss_data` in the Data list of the Environment window.

## View the `YRBSS` dataset

- Run the `View` command below within RStudio.
  - A new tab in the scripting window will appear with the `yrbss_data` dataset.
- Note that the code chunk specifies `#| eval: FALSE` so that this code is not run while rendering the .qmd file.

```{r}
#| eval: FALSE

View(yrbss_data)
```

# Summarizing data and `|>`

- Often we want to summarize columns of data
- Quantitative data and categorical/text data will be summarized differently
- We also introduce the [pipe symbol]{style="color:darkorange"} `|>`, which is used to string together commands

The [pipe is read "and then"]{style="color:darkorange"} and tells R the next function will be applied to the data on the left side of the pipe:

```{r}
yrbss_data |>
  head() # shows top rows
```

# Summarizing quantitative data

- The **`rstatix` package** has a useful function called `get_summary_stats()` for quick statistics of quantitative data

```{r}
# library(rstatix)

yrbss_data |> 
  get_summary_stats(
    bmi, stweight,
    type = "common"
    ) 
```

- The `type` options are
  - `type = c("full", "common", "robust", "five_number", "mean_sd", "mean_se", "mean_ci", "median_iqr", "median_mad", "quantile", "mean", "median", "min", "max")`

## Stratified summaries (and more on `|>`)

- Now let's **summarize the BMI and weight variables by a strata subgroup**.
- We do this with the `group_by()` function
- Notice that we now have two `|>`:
  - `data |> group_by(...) |> get_summary_stats(...)`
  - With the pipe symbol `|>` we can **string together many commands** in the order we want them applied. Example:
    - first tell R what dataset to use,

    - then have R group (stratify) the data by `sex`,

    - and then calculate summary statistics

```{r}
yrbss_data |> 
  group_by(sex) |>
  get_summary_stats(
    bmi, stweight,
    type = "common"
    )
```

## Stratify by 2 variables

- We can specify more than one variable within `group_by()` to stratify the summaries by more variables

```{r}
yrbss_data |> 
  group_by(sex, grade) |>
  get_summary_stats(
    bmi, stweight,
    type = "common"
    )
```

## Prettier table of output

- Below we add on `gt()` at the end to make the table prettier.
  - Make sure to check this out in the html output.
- The `gt()` command is from the `gt` [package](https://gt.rstudio.com/)

```{r}
yrbss_data |> 
  group_by(sex, grade) |>
  get_summary_stats(
    bmi, stweight,
    type = "common"
    ) |>
  gt()
```

## Advanced examples using `summarize` function

Below are links to data summary examples in our previous OCTRI-BERD workshop on *Introduction to R and RStudio for Exploratory Data Analysis: Part 2* using the function `summarize()`:

- [Summarize examples: Slides 6 - 12](https://jminnier-berd-r-courses.netlify.app/01-intro-r-eda/01_intro_r_eda_part2.html#6)
- All slides:
  - html slides: [bit.ly/berd_intro_part2](http://bit.ly/berd_intro_part2)
  - pdf slides: [bit.ly/berd_intro_part2_pdf](http://bit.ly/berd_intro_part2_pdf)


# Two pipes in R: `|>` vs `%>%`

## Two pipes in R: `|>` vs `%>%`

- Both pipes are used to write code from **left to right** 
    -   by passing the result of one step into the next. 

### Background

- The newer **base R pipe `|>`** was introduced in 2021 with the release of R 4.1.0.
- The original **magrittr pipe `%>%`** 
    -   comes from the {magrittr} package and 
    -   became popular through the tidyverse.


-   Using the **base R pipe `|>`** does *not* require loading the {tidyverse} package (which includes {magrittr}),
    -   while using the **magrittr pipe `%>%`** does require {tidyverse}.

### Similarities

-   They work the same when the left-hand side becomes the **first argument** of the next function.
-   Example: 


Using `|>`
```{r}
yrbss_data |> dim()
```

Using `%>%`
```{r}
yrbss_data %>% dim()
```



### Differences

*When do `|>` and `%>%` behave differently?*


#### [Simple example]{style="color:purple"}


- `|>` is **stricter** and expects an actual function call on the right

```{r}
yrbss_data |> dim()
```

```{r}
#| error: true
yrbss_data |> dim
```

-   `yrbss_data |> dim()` works, but  `yrbss_data |> dim` does not: 

With `%>%`, we can omit the `()` after the function name:
```{r}
yrbss_data %>% dim()
yrbss_data %>% dim
```


#### [Advanced example]{style="color:purple"}

-   Code below runs a linear regression model with dependent variable `bmi` and independent variable `age`.

- Using `%>%`: 
    -   note that for the data argument a `.` is used as the placeholder for the dataset `yrbss_data` since it's already being "fed into" the model.

```{r}
yrbss_data %>% lm(bmi ~ age, data = .)
```

- Using `|>`: 
-   Using a `.` as as a placeholder for the data does not work with the newer base R `|>`:

```{r}
#| error: true
yrbss_data |>  lm(bmi ~ age, data = .)
```

-   Starting with R 4.3.0, we can use a placeholder but it's an underscore`_`: 

```{r}
yrbss_data |>  lm(bmi ~ age, data = _)
```

#### [More advanced examples]{style="color:purple"}

See `Learn more` section below for resources with more examples.



### For this workshop

We will use `|>` because 

-   it is built into R, 
-   requires no extra package, and 
-   works well for most tidyverse code you will write as beginners

Our older workshops and many reference materials use `%>%`. 


### Learn more

-   [Differences between the base R and magrittr pipes](https://tidyverse.org/blog/2023/04/base-vs-magrittr-pipe/){target="_blank"} - Tidyverse blog post by Hadley Wickham
- [Discussion of differences on Stack Overflow](https://stackoverflow.com/questions/67633022/what-are-the-differences-between-rs-native-pipe-and-the-magrittr-pipe){target="_blank"}
- [R 4.1.0 pipe overview](https://www.jumpingrivers.com/blog/new-features-r410-pipe-anonymous-functions/){target="_blank"}
-   [Comparing pipes: Base-R |> vs {magrittr} %>%](https://albert-rapp.de/posts/31_pipes_compared/31_pipes_compared){target="_blank"} 
    -   this resource includes examples and 
    -   also how to change the "default" pipe in RStudio from `%>%` to `|>` (keyboard shortcut will use `|>`)


# Summarizing categorical data: frequency tables

- We will be creating frequency tables using the **`tabyl` function from the `janitor` package** (loaded above).

## Frequency table for one variable

Default table:

```{r}
yrbss_data |> 
  tabyl(age)  
```

Make the output prettier using `adorn_` options and adding on `gt()`:

With adornments:

```{r}
yrbss_data |> 
  tabyl(age)   |>
  adorn_totals("row") |> 
  adorn_pct_formatting(digits=2)
```

With adornments & `gt()`:

```{r}
yrbss_data |> 
  tabyl(age)   |>
  adorn_totals("row") |> 
  adorn_pct_formatting(digits=2) |> 
  gt() # from the gt package
```

## Contingency tables (two-way tables)

- **Contingency tables** summarize data for two categorical variables
  - with each value in the table representing the number of times\
    a particular combination of outcomes occurs
- **Row & column totals** added with `adorn_totals` option
  - Can choose to include row and/or column totals

```{r}
#| output-location: column

# ages vs. grades:

yrbss_data |> 
  tabyl(age, grade) |>    
  adorn_totals(c("row", "col")) |> 
  adorn_title(placement = "combined") |> 
  gt()
```

## Contingency tables with percentages

- How are the percentages calculated in the code below?
- What else was added to the table, other than the percentages?

```{r}
yrbss_data |> 
  tabyl(age, grade) |>    
  adorn_totals(c("row")) |>
  adorn_percentages("row") |>   
  adorn_pct_formatting(digits=0) |>    
  adorn_ns() |> 
  gt()
```

## Stratified (3-way) tables

We can also add a third variable, which creates 2x2 tables stratified by the third variable:

```{r}
yrbss_data |> 
  tabyl(age, grade, sex)
```

*Make sure to scroll down in the output box above to see all of the output*

## More `tabyl` and `adorn_` options

See the [`tabyl`s vignette](https://cran.r-project.org/web/packages/janitor/vignettes/tabyls.html) for more `tabyl` and `adorn_` options:

<https://cran.r-project.org/web/packages/janitor/vignettes/tabyls.html>

# Practice time! (part 2)

1.  Create a contingency table for `race4` and `race7`, with both percentages and counts. What is the difference between these two variables?

```{r}

```

2.  Create a contingency table for `race4` and `race7`, with both percentages and counts, stratified by `grade`.

```{r}

```

3.  Calculate the minimum, median, and maximum BMI and weights of participants in the YRBSS data.

```{r}

```

4.  Calculate the minimum, median, and maximum BMI and weights of the participants in the YRBSS data, now stratified by age.

```{r}

```

# Data visualization

- This is a *very brief* introduction to creating plots using the `ggplot2` package.
  - The `ggplot2` package gets loaded with the `tidyverse` and thus we do not need to load it separately.

# Visualizing quantitative variables

- What is the distribution shape of a quantitative variable?
- Does the distribution shape of a quantitative variable change when stratified by a categorical variable?
- Is there an association between two quantitative variables?

## Histograms

```{r}
ggplot(data = yrbss_data,    # specify the dataset
       # aes = aesthetics: specify variables being used
       aes(x = bmi)) +  
  geom_histogram() # type of plot
```

## Boxplots

- Recall that the whiskers' ends are the minimum and maximum values
- The box is from the first to third quartiles, with the line inside the box at the median

**BMI**: horizontal boxplot (specify `x = ...`)

```{r}
#| fig.height: 3
#| fig.width: 5

ggplot(data = yrbss_data,   # dataset
  # aes: specify vars being used
       aes(x = bmi)) +   
  geom_boxplot()   # type of plot
```

**Weight**: vertical boxplot (specify `y = ...`)

```{r}
#| fig.height: 3
#| fig.width: 5

ggplot(data = yrbss_data, 
# use y instead of x in aes for vertical:
  aes(y = stweight)) + 
  geom_boxplot() 
```

## Stratified boxplots (plus axis labels & title!)

```{r}
ggplot(data = yrbss_data, 
       aes(
         # horizontal boxplots for BMI (x=):
         x = bmi,
         # stratified vertically (y=) by age:
         y = age
         )) + 
  geom_boxplot() + 
  # Add on labels and a title:
  labs(x = "BMI", 
       y = "Age", 
       title =  "BMI distributions by age category"
       )  
```

## Add color using `fill`

- Also remove the `NA` category with `drop_na()` (we will talk about this in later slides)
- What's the difference in use of `fill=` here?

```{r}
#| fig.height: 5
#| fig.width: 7
ggplot(
  data = yrbss_data |> drop_na(age), #NEW!
  aes(
    x = bmi,
    y = age,
    fill = age  # NEW!
  )) + 
  geom_boxplot() +
  # remove legend:
  theme(legend.position="none") 
```

```{r}
#| fig.height: 5
#| fig.width: 7
ggplot(
  data = yrbss_data |> drop_na(age), #NEW!
  aes(
    x = bmi,
    y = age,
    fill = sex  # NEW!
  )) + 
  geom_boxplot() 
```

## Scatterplots

```{r}
ggplot(data = yrbss_data, # specify the dataset
       # aes = aesthetics: specify variables being used
       aes(x = stweight,       
           y = bmi
           )) +  
  geom_point()    # type of plot     
```

## Scatterplot with color-coded dots

```{r}
ggplot(data = yrbss_data, 
       # aes = aesthetics: specify variables being used
       aes(x = stweight,      
           y = bmi,  
           color = sex,
           )) +   
  geom_point(alpha = .2) # alpha controls transparency of the points
```

## Color vs. fill vs. alpha

- `color` (or `colour`) is used for lines and outlines

- `fill` is used for interiors

- Note: geom_point() shape default is a point with only an outline and no interior -\> *we used color and not fill*

- `alpha` specifies the opacity / transparency

  - Values range from 0 to 1
  - lower alphas are more transparent

- See ggplot webpage [Colour related aesthetics: colour, fill, and alpha](https://ggplot2.tidyverse.org/reference/aes_colour_fill_alpha.html) for more information

# Visualizing categorical variables

## Barplots

By default, barplots display counts (frequencies) on the vertical axis:

```{r}
#| fig.width: 10
ggplot(data = yrbss_data, 
       aes(x = age)) +
  geom_bar() + # plot type 
  # angle the x-axis labels so that they are not overlapping
  theme(axis.text.x = element_text(angle = -45, hjust = 0))
```

## Barplots with proportions

- To show proportions, the code is more complicated.

```{r}
#| fig.width: 10
ggplot(data = yrbss_data, 
       aes(x = age)) +
  # specify aesthetics within the barplot to show proportions
  geom_bar(aes(y = after_stat(prop), group = 1)) + 
  
  # Next line converts y-axis labels to percentages instead of proportions
  scale_y_continuous(labels =  scales::percent_format()) +
  
  # "dodge" the x-axis labels
  scale_x_discrete(guide = guide_axis(n.dodge = 2))
```

## Barplots with 2 variables: segmented bar plots

Stacked bars showing counts:

```{r}
#| fig.width: 10
ggplot(data = yrbss_data, 
       aes(x = age,
           # specify variable for colors 
           fill = grade)) + 
  geom_bar()
```

For proportions within each bar:\
(added `position = "fill"` within `geom_bar()`)

```{r}
#| fig.width: 10
ggplot(data = yrbss_data, 
       aes(x = age,
           # specify variable for colors
           fill = grade)) + 
  geom_bar(position = "fill")
```

## Side-by-side barplots

```{r}
#| fig.width: 10
ggplot(data = yrbss_data, 
       aes(x = age,
           # specify variable for colors 
           fill = grade)) + 
  geom_bar(position = "dodge") +
  # angle the x-axis labels so that they are not overlapping
  theme(axis.text.x = element_text(angle = -15, hjust = 0))
```

## Saving a ggplot as an image file

First, create a plot and save it as an R object:

```{r}
myplot <- ggplot(data = yrbss_data, 
       aes(x = age,
           # specify variable for colors 
           fill = grade)) + 
  geom_bar() +
  # angle the x-axis labels so that they are not overlapping
  theme(axis.text.x = element_text(angle = -15, hjust = 0))

myplot
```

Save the plot as a pdf (or "jpeg", "tiff", "png", etc.)

```{r}
ggsave(
  filename = "output/myplot.pdf",
  plot = myplot,
  device = "pdf",
  width = 5,
  height = 3,
  units = "in",
  dpi = 200)  
```

Can specify the dpi when saving, many journals have dpi requirements.

See `ggsave` webpage for more details: <https://ggplot2.tidyverse.org/reference/ggsave.html>

# Missing data summary & visualization

- I *highly recommend* the tutorial: **Exploring missing values in `naniar`** <https://allisonhorst.shinyapps.io/missingexplorer/>
- These slides have some highlights of tools presented in the tutorial

Load the `naniar` package if haven't already:

```{r}
library(naniar)
```

## Missingness statistics

Total number of missing values in dataset:

```{r}
n_miss(yrbss_data)
```

Proportion of missing values in dataset:

```{r}
prop_miss(yrbss_data)
```

Missingness by each variable:

```{r}
miss_var_summary(yrbss_data) |> gt()
```

## Visualizing missing data across dataset

- This plot helps visualize whether
  - there are variables with a lot of missing values and/or
  - participants with a lot of missing values
- Each row is an observation
- Columns are for the different variables in the dataset
- Black = missing, grey = not missing

```{r}
vis_miss(yrbss_data) 
```

## Missingness heatmap by group

- This plot is useful in comparing missingness rates stratified by a variable
- Specify the variable to stratify the missingness heatmap by in the `fct =` option

Example: Percent missingness stratified by age group

```{r}
gg_miss_fct(yrbss_data, fct = age) 
```

## Missingness intersections (UpSet plot)

- UpSet plots are great for finding sets of variables that have a lot of missingess.

```{r}
gg_miss_upset(yrbss_data, nsets = 10) 
```

# Practice time! (part 3)

1.  Create a histogram for the `stweight` variable. Describe the distribution shape.

```{r}

```

2.  What happens if you change `x =` to `y =` in the histogram code?

```{r}

```

3.  Create boxplots of `stweight` stratified by sex.

```{r}

```

4.  Create a barplot of race4 categories with the y axis showing percentages.

```{r}

```

5.  Visualize the missing data across the YRBSS subset dataset we created `toy_data` using `vis_miss()`. What missingness patterns do you see?

```{r}

```

# Data wrangling

## **Sorting** data frames using `arrange()`

`arrange()` is a function that lets us **sort** a `data.frame` by a specified variable.

- By default, it sorts in *ascending* order:

```{r}
yrbss_data |> 
  arrange(record)
```

- To sort by descending order, you need to wrap the variable in the `desc()` function:

```{r}
yrbss_data |> 
  arrange(desc(record))
```

## Sorting data frames by multiple variables

- You can also arrange by multiple variables.
  - The code below sorts by `age` *first*, and then within `age`, we sort by `record`.

```{r}
yrbss_data |> 
  arrange(age, record)
```

- Note that order of variables in `arrange()` matters!

```{r}
yrbss_data |> 
  arrange(record, age)
```

## `arrange()` after `count()`-ing categorical data

- Another example of when `arrange()` is useful

- We've used `tabyl()` to see what values exist in a categorical variable.

- We can also use `count()` to count all of the unique values for a categorical variable.

- This will also show `NA` category, which is the special missing variable type:

```{r}
yrbss_data |>
  count(age)
```

- Sometimes it helps to arrange by that category if you have many categories to view.
- This can also alert us if there are categorical coding mistakes (such as misspellings) in our data.

```{r}
yrbss_data |>
  count(age) |>
  arrange(n)
```

## Saving changes we make to a dataframe

- Note that we have not changed the order of the `yrbss_data` dataframe after arranging:

```{r}
head(yrbss_data)
```

- We need to *save our work* by re-assigning the sorted/arranged data frame as `yrbss_data` with the `<-` operator:

```{r}
yrbss_data <- yrbss_data |> arrange(desc(record))

yrbss_data
```

# Subsetting data

# Subset rows: `filter()`ing data

## `filter()`ing quantitative data

`filter()` is an extremely powerful function. It lets us **subset our data** according to specific criteria.

- Let's filter on a `numeric` variable, `bmi`:

***Note the double equal signs!*** `==` is a *test* of whether two things are equal, it is not an assignment (i.e. `x = 5` assigns the value 5 to the object `x`).

```{r}
yrbss_data |>
  filter(bmi > 40)

yrbss_data |>
  filter(bmi == 13.9)

yrbss_data |>
  filter(bmi == 13.9229)
```

## `filter()`ing cateogrical data

- To filter on a categorical variable, we need to know the values (levels) of the variable:

```{r}
yrbss_data |> tabyl(grade)
```

[***Note the double equal signs!***]{style="color:darkorange"} `==` is a *test* of whether two things are equal, it is not an assignment (i.e. `x = 5` assigns the value 5 to the object `x`).

```{r}
yrbss_data |>
  filter(grade == "10th") #We must include " " around the value for categorical variables
```

- If we want everything except 10th grade, we use the ["not equal" sign]{style="color:darkorange"} `!=`

```{r}
yrbss_data |>
  filter(grade != "10th")
```

## Filtering requires a little logic

- We can chain multiple criteria using the [`&` (AND)]{style="color:darkorange"} or [`|` (OR) operators]{style="color:darkorange"}.
  - But we need to review a little logic before we do this.
- If we wanted to return participants who identified as
  - `Female` **and** are in 10th grade,
  - we would use an `&` to chain these criteria together:

```         
sex == "Female" & grade == "10th"
```

- Note that *SPELLING AND CASE MATTER*!
  - You must write the name of the column exactly and
  - also the value (i.e. "Female" not "female") exactly.

## Filtering with `&` (AND) criteria

```{r}
yrbss_data |>
  filter(sex == "Female" & 
           grade == "10th")

```

- Note that we could also use the comma in this case:

```{r}
yrbss_data |>
  filter(sex == "Female",    # comma is equivalent to & within filter()
         grade == "10th")

```

## Filtering with `|` (OR) criteria

If we wanted patients who were

`Female` [**or**]{style="color:darkorange"} in 10th grade

we would use a `|` to chain these criteria together.

```{r}
yrbss_data |>
  filter(sex == "Female" |
           grade == "10th")
```

- Think about it:
  - Which of the AND vs. OR code blocks will return a larger number of participants?

## Filtering out missing values using `drop_na()` (1/2)

- We may want to remove rows that have missing data, which are coded as `NA`.

- See the [`drop_na()`](https://tidyr.tidyverse.org/reference/drop_na.html) reference for examples (`?drop_na`).

- First look at rates of missingness using the `naniar` package:

```{r}
naniar::miss_var_summary(yrbss_data)
```

## Filtering out missing values using `drop_na()` (2/2)

*Pay attention to the number of rows for each of these results*

- Remove missingness in one column:

```{r}
yrbss_data |> drop_na(race7)
```

- Remove missingness in two columns:

```{r}
yrbss_data |> drop_na(race7, bmi)
```

- Remove all missingness:

```{r}
yrbss_data |> drop_na()
```

## A warning about `filter()`: check for NA

- Suppose we want to remove the `10th` grade category.
- However this column has missingness in it as we can see with `tabyl()`.
- When we try to filter using `!="10th"` we lose our `NAs`!
  - That isn't really expected.
- You need to account for this with an *or* `|` statement:

```{r}
yrbss_data |> 
  tabyl(grade)
```

Lost the `NA`'s in the filtering below!

```{r}
yrbss_data |> 
  filter(grade != "10th") |>
  tabyl(grade)
```

Updated code to keep the `NA`'s when filtering:

```{r}
yrbss_data |>
  # add an | (OR) statement 
  # to keep in the NA values
  filter((grade != "10th") |
           is.na(grade)) |>
  tabyl(grade)
```

# Subset columns: `select()`ing data

## Selecting columns using `select()`

- The `select()` function allows us to select variables or columns from our dataset:

```{r}
yrbss_data |>
  select(record, age, race4)

```

- If we want to select everything *except* one variable,
  - we can use a `-` in front of that variable.

```{r}
yrbss_data |>
  select(-race4)

```

## `tidyselect`

- There are ways to search column names and to select them.
- These are called the `tidyselect` helpers.
- You can see some examples here: <https://tidyselect.r-lib.org/reference/language.html>
- For instance, you can see how this might be useful, where we select all columns where the column name includes the word "race":

```{r}
yrbss_data |> 
  select(contains("race"))
```

- or select all columns starting with "r":

```{r}
yrbss_data |> 
  select(starts_with("r"))
```

- This is a useful shortcut (though use with caution, you need to know your column names and order):

```{r}
yrbss_data |> select(1:3)
```

- `select()` and `everything()` are a useful combination for some quick rearranging of columns:

```{r}
yrbss_data |> select(record, bmi, everything())
```

- Note there's also a function [`relocate()` to move around your column order](https://dplyr.tidyverse.org/reference/relocate.html).

## `tidyselect` mini-practice

- We can also select variables based on data types
- For instance, this use of `select(where(is.numeric))` selects all numeric columns
  - (`where()` is a helper function that returns columns where the inside function is TRUE,
  - in this case `is.numeric` returns the TRUE/FALSE)

```{r}
yrbss_data |> select(where(is.numeric))
```

- Try using `is.character` to select all character columns, as well as the column `bmi`.

```{r}
#| eval: FALSE
# change eval: TRUE above
yrbss_data |> select(____, _____)
```

## \*\* Learn more about `tidyselect` (BONUS - on your own)

- See some more examples in [this slide](https://jminnier-berd-r-courses.netlify.app/02-data-wrangling-tidyverse/02_data_wrangling_slides_part1.html#32) from a previous BERD workshop

- For more info and learning about `tidyselect`, please run this code in your console:

```{r}
#| eval: FALSE
# install remotes package
install.packages("remotes")

# use remotes to install this package from github
remotes::install_github("laderast/tidyowl")

# load tidyowl package
library(tidyowl)

# interactive tutorial
tidyowl::learn_tidyselect()
```

## The difference between `filter()` and `select()`

- One thing to keep in mind is that:
  - `filter()` works on [rows]{style="color:darkorange"} (think FILTER in Excel!), and
  - `select()` works on [columns]{style="color:darkorange"} (select your relevant variables)

Keep that in mind!

## Renaming with `rename()`

- You can rename columns with `select()` but it's a bit easier to do this with the `rename()` function.

- You just need to remember **`new_name = old_name`** ordering.

  - One equal sign will suffice!

```{r}
yrbss_data |> rename(id = record,
                  race_cat = race4,
                  weight = stweight)
```

*Note that [we did not save the changes above]{style="color:darkorange"}, and thus we haven't actually renamed the variables in the dataset `yrbss_data`!*

## Saving our results as a file

- Let's save some processed data in the `output/` directory.
- We'll save it as an `xlsx` file so that we can open it in Excel.

```{r}
# chain together several data transformation steps with |>

# use <- to assign the new data to a different name to avoid overwriting the original dataset

processed_data <- yrbss_data |>   
  select(-race7) |>
  rename(id = record,
         race_cat = race4,
         weight = stweight) |>
  filter(sex == "Female")

glimpse(processed_data)
```

[Save]{style="color:darkorange"} the `processed_data`:

```{r}
write_xlsx(x = processed_data,
                path = "output/processed_data.xlsx")
```

# Practice time! (part 4)

Perform some subsetting functions on the `yrbss_data` data:

1.  Use `filter()` to select people who are in 9th or 10th grade and who have BMI less than 25. Save the resulting data as `yrbss_filtered` tibble.

```{r}

```

2.  Use `rename()` to rename 2 of the column names of `yrbss_filtered` to something more meaningful.

```{r}

```

3.  Use `select()` to keep only the columns that denote record, sex, grade, bmi, and the 2 renamed columns.

```{r}

```

4.  Arrange the data by bmi, starting with the highest value and ending with the least value.

```{r}

```

5.  Save the resulting data set in the file `data/practice4_data.xslx`.

```{r}

```

# Make new variables: `mutate()`

## Using `mutate` to calculate a new variable based on other variables (1/3)

- One use for `mutate` is to do Excel type calculations using other columns in the data.
  - For instance, we might want to covert weight to kg, assuming `stweight` is in lbs
- Notice we use `=` inside mutate, not `==` or `<-`.
  - We are not making a logical operation, nor are we assigning something to an object name.

```{r}
yrbss_data |> 
    mutate(weight_kg = stweight/2.205) |>
    glimpse()
```

- Remember, since we haven't assigned this new "mutated" data frame to an object name, our work hasn't been saved.

## Using `mutate` to calculate a new variable based on other variables (2/3)

- Below, we save the new data frame/tibble as `yrbss_new`
  - to avoid overwriting our other data set.
  - Check out the environment tab for the new tibble.
- Notice also how we can mutate multiple variables at one time.
  - We can also mutate using multiple columns.
  - Here we have created two new weight variables in different scales
  - In the next *Practice time*, you will make a height variable from `bmi` and `weight_kg`

```{r}
yrbss_new <- yrbss_data |> 
    mutate(weight_kg = stweight/2.205,
           weight_g = weight_kg*1000)

glimpse(yrbss_new)
```

## Using `mutate` to calculate a new variable based on other variables (3/3)

Figure examining the newly created variables using `mutate`:

```{r}
ggplot(yrbss_new) +
  aes(x = stweight, 
      y = weight_kg) +
  geom_point()
```

## Brief aside: categorical data (`factor`s)

- One data type that we haven't yet looked at are `factor`s

  - [factors are how R represents **levels** of *categorical data*]{style="color:darkorange"}.

- For the most part, you can use `character` and `factors` interchangeably for categorical data.

- However, there is one main difference.

  - `factor`s define the permissible values in a vector with an argument called `levels`.
  - They also define the order in which these values are displayed.

```{r}
character_vector <- c("Dog", "Dog", "Cat", "Mouse")   # c() is the concatenate function
class(character_vector)
tabyl(character_vector)
```

- The default order of levels is in **alphanumeric** order.
- You can make a character vector into a factor vector by using the `factor()` function, and supplying an argument called `levels`.

## **Levels** of a `factor` variable

- The `levels` of a factor are the *permissible values* in a `factor`.

- The **order of the values in a factor variable** determine the order in which the values appear in tables and on the axes in a plot.

- You can control the order of the categories in a factor by specifying the order of the categories in the `levels` argument.

- You can find out the levels of the factor with the function `levels()` or `tabyl()`.

```{r}
factor_vector <- factor(character_vector)

class(factor_vector)

# what are the levels of this vector?
levels(factor_vector)

tabyl(factor_vector)
```

```{r}
# specify our own ordering of levels
factor_vector2 <- factor(character_vector, 
                        levels = c("Dog", "Cat", "Mouse"))

levels(factor_vector2)

tabyl(factor_vector2)
```

Being able to specify the ordering is the main reason to use `factors`, at least in plotting and calculating counts.

## Mini-practice

- Change the order of `factor_vector` to be "Mouse", "Cat", "Dog".

```{r}

```

- Verify that you did it correctly by calling `tabyl()` or `levels()`

```{r}
#| eval: FALSE

factor_vector <- factor(____, 
                        levels = _____)

tabyl(factor_vector)

levels(factor_vector)

```

## Using `mutate` to transform `character` variables into `factors`

- We can also use use `mutate()` to make a `character` variable a `factor`.

- Let's convert `age` from `character` into `factor`:

```{r}
yrbss_data |> 
  #create a factor variable
  mutate(age_fac = factor(age))
```

- An example with grade:

```{r}
yrbss_data |> 
  #create a factor variable
  mutate(grade = factor(grade)) 
```

- One thing to notice:
  - we are doing something called [*reassignment* in the grade example]{style="color:darkorange"}.
  - We're taking the previous values of our variable (`grade`),
    - doing something to it (making it a `factor`), and then reassigning the variable `grade` to our new set of values.

## `levels` of a `factor` variable revisited

- Remember when we wanted to decide the order of the categories of the factor?
  - We can assign that order using the `levels` argument in `factor()`.
  - We can check the ordering with `tabyl()`

```{r}
yrbss_data |>
  tabyl(grade)    # Order of grades is not correct
```

- Create grade_fac variable to be a factor version of grade with "correct" levels order:

```{r}
yrbss_data |> 
  mutate(grade_fac = factor(grade, 
                      levels = c("9th","10th","11th","12th"))) |>
  tabyl(grade_fac)
```

- Notice that the order of the grade categories is now correct

*Remember spelling matters!*

- If we try to use a factor level that doesn't exist ("10" vs "10th"),
  - we won't have any data in that category, and
  - the non-specified levels ("10") get set to `NA` missing.

```{r}
yrbss_data |> 
  mutate(grade_fac = factor(grade, 
                      levels = c("9th","10","11th","12th"))) |>
  tabyl(grade_fac)
```

# Practice time! (part 5)

1.  Create a new variable `height` that calculates height based on `bmi` and `weight_kg`. Note the formula for BMI is weight/height\^2 (weight is in kilograms and height is in meters). Create a scatterplot of `bmi` vs `height`.

```{r}

```

2.  Create a new factor variable `sex_fac` that orders the values of `sex` as Male, Female.

```{r}

```

# Make categorical variables with `case_when()`

## `case_when()`

- Currently the grade variable is a character vector. How do we change these to numeric?
- Or, how might we dichotomize a continuous variable such as bmi?
- Or, what if we want to combine categories?

First let's start with a simple binary 2 category variable.

- We need to follow the **basic pattern** for each of our categories inside `case_when()`:
  - [*condition* \~ *category name*]{style="color:darkorange"}
  - The *left* side of the `~` is where we can specify how we *define* the category based on our column variable names.
  - The *right* side of the `~` is where we can specify the *category name* (as a character).
  - Separate each "case" with a comma.

## `case_when()` example 1

- In the example below, we are converting `sex` to a character variable called `female`:
  - `sex == "Female"` tests whether the column `sex` is Female, which is how we define the category 1
    - 1 is the new value, and this is the right side of `~`
  - `sex == "Male"` tests whether the column `sex` is Male, which is how we define the category 0
    - 0 is the new value, and this is the right side of `~`
- Note how `NA` is handled here:

```{r}
yrbss_data |>
  mutate(female = case_when(
    sex == "Female" ~ 1, # do not forget this comma!
    sex == "Male" ~ 0
  )) |>
  tabyl(female) # look at the categories
```

## `case_when()` example 2

- In the example below, we are dichotomizing the continuous variable bmi into two categories
- We need to do this for each level in our category:

[*left side* defines the condition *\~* *right side* names the category]{style="color:darkorange"}

```{r}
yrbss_data |>
  mutate(
    bmi30 = case_when(
      bmi < 30 ~ "<30", # don't forget a comma between cases
      bmi >= 30 ~ "30+"
    )
  ) |>
  tabyl(bmi30)
    
```

## `case_when()` example 3

Next we recode bmi to have 3 levels:

- `<20` (not including exactly 20)
- `20-30` (not including exactly 30)
- `30+`

Notice the middle category we use an & statement similar to how we used logic in filter

```{r }
yrbss_data |>
  mutate(
    bmi3cat = case_when(
      bmi < 20 ~ "<20",
      (bmi >=20) & (bmi < 30) ~ "20-30",
      bmi >= 30 ~ "30+"
    )
  ) |>
  mutate(bmi3cat = factor(bmi3cat)) |>   # make it factor after creating categories
  tabyl(bmi3cat)
```

# Practice time! (part 6)

1.  Create a new category variable called `grade_num` in `yrbss_data` that converts `grade` to numeric.

```{r}

```

2.  Create a new variable called `race_cat` that relabels `race4` to have shorter race category names: W (White), H/L (Hispanic/Latino), B/AA (Black or African American), Other (All other races). Make `race_cat` a factor variable with levels ordered by the size of the group in descending order.

```{r}

```

3.  Create a `tabyl` of the new variable `race_cat` and `race4` to check your work, including the ordering of the factors.

```{r}

```

# Try what we've learned with making figures using `ggplot2`

## Recall boxplots `geom_boxplot()`

- Boxplots can be used to compare the distribution of a quantitative variable among categories.
  - We need a categorical variable (here, our `x` axis)
  - and a continuous numeric variable, shown here on our `y` axis.

```{r}
#| output-location: column

# creating a boxplot

ggplot(yrbss_data) +
  aes(x = grade, 
      y = bmi) +
  geom_boxplot()
```

- The main differences from the scatterplots we created earlier are the `geom` type and the variables plotted.

- We can change the color similarly to scatterplots.

  - However, we map to `fill` and not `color` if we want to fill in the box with color:

```{r}
#| output-location: column

# adding color
ggplot(yrbss_data) +
  aes(x = grade, 
      y = bmi,
      fill = grade) +
  geom_boxplot()
```

## Factor levels when plotting

- Perhaps we want to **change the order of the `grade` levels** in our plot.
- This is where factor levels are useful:

```{r}
yrbss_new <- yrbss_data |> 
  mutate(grade = factor(grade, levels = c("9th", "10th", "11th", "12th"))) |>
  drop_na(grade)

ggplot(yrbss_new) +
  aes(x = grade, 
      y = bmi,
      fill = grade) +
  geom_boxplot()
```

## Faceting our boxplot with `facet_wrap()`

- One of the most powerful ways to change a visualization is by *faceting*.
  - We can make multiple plots stratifying by a categorical variable.
- To do this, we have to add the `facet_wrap()` command to our plot.
  - We need to specify the variable to `facet_wrap` - `race4` by using the `vars()` function to specify it as a variable.

```{r}
#| fig.height: 8

# facet_wrap:
ggplot(yrbss_new) +
  aes(x = bmi, 
      y = grade,
      fill = grade) +
  geom_boxplot() +
  facet_wrap(vars(factor(race4)))  #NEW!
```

- *Don't try to facet on a continous `numeric` variable - it won't work.*

- Don't forget to look at the help documentation (e.g., `?facet_wrap`) to learn more about additional ways to customize your plots!

## Faceting our boxplot with `facet_grid()`

- Another way to facet our plots is with `facet_grid()`,
  - which lets us select the rows and columns based on categorical variables:

```{r}
#| fig.height: 8
#| fig.width: 12

ggplot(yrbss_new) +
  aes(x = bmi, 
      y = sex,
      fill = sex) +
  geom_boxplot() +
  facet_grid(rows = vars(grade),  #NEW!
             cols = vars(race4))
```

## Mini-practice

Facet the boxplot below by `age` and `grade`, don't forget the `vars()`:

```{r}
#| eval: FALSE
# change eval: TRUE above

ggplot(yrbss_new) +
  
  aes(x = bmi, 
      y = sex,
      fill = sex) +
  
  geom_boxplot() + 
  
  facet_grid(______)
```

## Back to `mutate`: using `mutate` to replace missing values with `replace_na`

- Sometimes we want to fill in missing values with a certain value.
- We can use the `replace_na()` function inside of `mutate()` to specify this.
- For example, suppose if the bmi variable is missing,
  - we want to replace it with the average bmi across the entire data set.

```{r}
yrbss_new <- yrbss_data |>
  mutate(bmi_filled =
           replace_na(bmi,
                      mean(bmi, na.rm = TRUE)))           
yrbss_new |> 
  select(contains("bmi")) |> 
  tail()
```

- We can also fill in missing values for character vectors
  - Here we replace missingness in the `grade` column with `unknown`:

```{r}
yrbss_new <- yrbss_new |>
    mutate(gradefilled = replace_na(grade, "unknown"))

yrbss_new |> 
  tabyl(gradefilled, grade)
```

# Practice time! (part 7)

1.  Create a barplot of `race4`, faceted by `sex`.

```{r}

```

# Useful keyboard shortcuts

[Full list of keyboard shortcuts](https://support.posit.co/hc/en-us/articles/200711853-Keyboard-Shortcuts)

| action                              | mac                | windows/linux   |
|-------------------------------------|--------------------|-----------------|
| Run code in qmd (or script)         | cmd + enter        | ctrl + enter    |
| `<-`                                | option + -         | alt + -         |
| interrupt currently running command | esc                | esc             |
| keyboard shortcut help              | option + shift + k | alt + shift + k |

 

**Practice:** Try typing code below in your qmd (with shortcut) and evaluating it (with shortcut):

```{r}
#| eval: false
y <- 5
y
```
