---
title: "Practice Solutions"
subtitle: "Getting Started in R and Rstudio"
institute: "OHSU OCTRI-BERD"
author: "Jessica Minnier, Meike Niederhausen"
date: "04/30/2026"
format: 
  html:
    link-external-newwindow: true
    toc: true
    toc_float:
      collapsed: false
    number-sections: false
    number-depth: 4
    theme: minty
    embed-resources: true
    html-math-method: mathjax
execute:
  echo: true
editor: visual
editor_options: 
  chunk_output_type: inline
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

# Load packages 
library(tidyverse)
library(writexl)
library(rstatix)
library(janitor)
library(naniar)
library(gt)
library(gtsummary)
library(readxl)  # part of tidyverse, but not the tidyverse core
```

# Part 1

Open the Excel file `yrbss_demo.xlsx` in the *data* *folder* to familiarize yourself with it.

Use R code to answer the following questions.

## 1

1.  Load the dataset `yrbss_demo.xlsx`. Make sure the name of the loaded data is `yrbss_data`.

```{r}
# library(readxl)
# note we are saving the data from the Excel file in our environment 
# with the object name yrbss_data
yrbss_data <- read_excel("data/yrbss_demo.xlsx")
```

## 2

2.  What are the column names?

```{r}
names(yrbss_data)

# can also use glimpse:
glimpse(yrbss_data) 
```

## 3

3.  How many rows and columns are in `yrbss_data`?

```{r}
nrow(yrbss_data)

ncol(yrbss_data)

# both rows & columns:
dim(yrbss_data)
```

## 4

4.  Which variables are numeric (quantitative)?

```{r}
glimpse(yrbss_data) 
```

- Record, bmi, and stweight are numeric, although record actually serves as a categorical variable.
- Interestingly age and grade are not.

## 5

5.  Which variables are character?

```{r}
glimpse(yrbss_data) 
```

- Age, sex, grade, race4, and race7 are character variables.

# Part 2

## 1

1.  Create a contingency table for `race4` and `race7`, with both percentages and counts. What is the difference between these two variables?

```{r}
yrbss_data %>%
  tabyl(race7, race4) %>%
  adorn_percentages() %>%
  adorn_pct_formatting() %>%
  adorn_ns() %>%
  adorn_title(placement = "combined") %>%
  gt()
```

## 2

2.  Create a contingency table for `race4` and `race7`, with both percentages and counts, stratified by `grade`.

```{r}
yrbss_data %>%
  tabyl(race7, race4, grade) %>%
  adorn_percentages() %>%
  adorn_pct_formatting() %>%
  adorn_ns() %>%
  adorn_title(placement = "combined") # gt doesn't work here
```

## 3

3.  Calculate the minimum, median, and maximum BMI and weights of participants in the YRBSS data.

There are a couple of options to do this. I chose the five number summary option.

```{r}
yrbss_data %>% 
  get_summary_stats(
    bmi, stweight,
    type = c("five_number")
    ) %>% 
  gt()
```

## 4

4.  Calculate the minimum, median, and maximum BMI and weights of the participants in the YRBSS data, now stratified by age.

```{r}
yrbss_data %>% 
  group_by(age) %>% 
  get_summary_stats(
    bmi, stweight,
    type = c("five_number")
    ) %>% 
  gt()
```

# Part 3

## 1

1.  Create a histogram for the `stweight` variable. Describe the distribution shape.

```{r}
ggplot(yrbss_data, aes(x = stweight)) + 
  geom_histogram()
```

The student weights have a skewed right distribution, with a long tail for higher weights.

## 2

2.  What happens if you change `x =` to `y =` in the histogram code?

The axes are flipped and we get a horizontal histogram.

```{r}
ggplot(yrbss_data, aes(y = stweight)) + 
  geom_histogram() # axes are flipped
```

## 3

3.  Create boxplots of `stweight` stratified by sex.

The weights of males are slightly greater than for females.

```{r}
ggplot(yrbss_data, 
       aes(x = stweight, y = sex)) + 
  geom_boxplot() 
```

## 4

4.  Create a barplot of race4 categories with the y axis showing percentages.

```{r}
ggplot(data = yrbss_data, 
       aes(x = race4)) +
  # specify aesthetics within the barplot to show proportions
  geom_bar(aes(y = after_stat(prop), group = 1)) +
  # Next line converts y-axis labels to percentages instead of proportions
  scale_y_continuous(labels =  scales::percent_format())   +
  labs(y = "Percent") # default label is prop
```

## 5

5.  Visualize the missingness patterns of the `toy_data` YRBSS subset using `vis_miss()`. What missingness patterns do you see?

```{r}
toy_data <- read_excel("data/toy_data.xlsx", na = "NA")

vis_miss(toy_data)
```

We see a lot of missingness in the `text_while_driving` variable (about half missing), and some missingness for `smoked_ever`, `bullied_past_12mo`, and `race4`.

# Part 4

Perform some subsetting functions on the `yrbss_data` data:

## 1

1.  Use `filter()` to select people who are in 9th or 10th grade and who have BMI less than 25. Save the resulting data as `yrbss_filtered` tibble.

```{r}
# first let's see what the categories look like
yrbss_data %>%
  tabyl(grade)

# now we filter with the "or" operator | on grade and then the "and" operator or comma on bmi
yrbss_filtered <- yrbss_data %>%
  filter(grade == "9th" | grade == "10th", 
         bmi < 25
         )

# check this worked
nrow(yrbss_filtered)
yrbss_filtered %>% 
  group_by(grade) %>% 
  get_summary_stats(
    bmi,
    type = c("common")
    ) %>% 
  gt()

# Note that output only includes 9th & 10th grade
# and max BMI is < 25
```

## 2

2.  Use `rename()` to rename 2 of the column names of `yrbss_filtered` to something more meaningful.

```{r}
names(yrbss_filtered)

yrbss_filtered2 <- yrbss_filtered %>% 
  rename(ID = record, 
         weight = stweight)

names(yrbss_filtered2)
```

## 3

3.  Use `select()` to keep only the columns that denote record, sex, grade, bmi, and the 2 renamed columns.

```{r}
yrbss_filtered3 <- yrbss_filtered2 %>%
  select(ID, sex, grade, bmi, weight)

# check:
glimpse(yrbss_filtered3)
```

## 4

4.  Arrange the data by bmi, starting with the highest value and ending with the least value.

```{r}
yrbss_filtered3 <- yrbss_filtered3 %>%
  arrange(desc(bmi))

yrbss_filtered3
```

## 5

5.  Save the resulting data set in the file `data/practice4_data.xslx`.

```{r}
write_xlsx(yrbss_filtered3, path = "data/practice4_data.xlsx")
```

# Part 5

*Note: code below is using `yrbss_data` instead of `yrbss_new`.*

## 1

1.  Create a new variable `height` that calculates height based on `bmi` and `weight_kg`. Note the formula for BMI is weight/height\^2 (weight is in kilograms and height is in meters). Create a scatterplot of `bmi` vs `height`.

```{r}
yrbss_data <- yrbss_data %>% 
  mutate(height = sqrt(stweight/bmi))

ggplot(yrbss_data, aes(x=height, y = bmi)) + 
  geom_point()
```

## 2

2.  Create a new factor variable `sex_fac` that orders the values of `sex` as Male, Female.

```{r}
yrbss_data <- yrbss_data %>%
  mutate(sex_fac = factor(sex,
                          levels = c("Male", "Female")))
         
# check:
glimpse(yrbss_data)

yrbss_data %>% tabyl(sex_fac, sex)
```

# Part 6

*Note: code below is using `yrbss_data` instead of `yrbss_new`.*

## 1

1.  Create a new variable called `grade_num` in `yrbss_data` that converts `grade` to numeric.

```{r}
yrbss_data %>% tabyl(grade)

yrbss_data <- yrbss_data %>% 
  mutate(grade_num = 
           case_when(
             grade == "10th" ~ 10, 
             grade == "11th" ~ 11, 
             grade == "12th" ~ 12, 
             grade == "9th" ~ 9))

# check:
yrbss_data %>% tabyl(grade, grade_num)

glimpse(yrbss_data)

```

## 2

2.  Create a new variable called `race_cat` that relabels `race4` to have shorter race category names: W (White), H/L (Hispanic/Latino), B/AA (Black or African American), Other (All other races). Make `race_cat` a factor variable with levels ordered by the size of the group in descending order.

```{r}
yrbss_data %>% tabyl(race4) %>% arrange(desc(n))

yrbss_data <- yrbss_data %>%
  mutate(
    race_cat = case_when(
      race4 == "White" ~ "W",
      race4 == "Hispanic/Latino" ~ "H/L",
      race4 == "Black or African American" ~ "B/AA",
      race4 == "All other races" ~ "Other"
      ),
    race_cat = factor(race_cat, levels = c("W", "Other", "H/L", "B/AA"))
    )

```

## 3

3.  Create a `tabyl` of the new variable `race_cat` and `race4` to check your work, including the ordering of the factors.

```{r}
# Check:
yrbss_data %>% tabyl(race4, race_cat) |> adorn_title()
```

# Part 7

*Note: code below is using `yrbss_data` instead of `yrbss_new`.*

## 1

1.  Create a barplot of `race4`, faceted by `sex`.

```{r}
ggplot(yrbss_data, aes(x=race4)) + 
  geom_bar() +
  facet_wrap(vars(sex)) +
  theme(axis.text.x = element_text(angle = -45, hjust = 0))
```
