---
title: "Practice Solutions"
subtitle: "Data Wrangling & Summarizing in R"
institute: "NW NARCH"
author: "Jessica Minnier"
date: "06/18/2026"
format: 
  html:
    link-external-newwindow: true
    toc: true
    toc_float:
      collapsed: false
    number-sections: false
    number-depth: 4
    theme: minty
    embed-resources: true
    html-math-method: mathjax
execute:
  echo: true
editor: visual
editor_options: 
  chunk_output_type: inline
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

# Load packages 
library(tidyverse)
library(writexl)
library(rstatix)
library(janitor)
library(naniar)
library(gt)
library(gtsummary)
library(readxl)  # part of tidyverse, but not the tidyverse core
```

# Part 1

Open the Excel file `yrbss_demo.xlsx` in the *data* *folder* to familiarize yourself with it.

Use R code to answer the following questions.

## 1

1.  Load the dataset `yrbss_demo.xlsx`. Make sure the name of the loaded data is `yrbss_data`.

```{r}
# library(readxl)
# note we are saving the data from the Excel file in our environment 
# with the object name yrbss_data
yrbss_data <- read_excel("data/yrbss_demo.xlsx")
```

## 2

2.  What are the column names?

```{r}
names(yrbss_data)

# can also use glimpse:
glimpse(yrbss_data) 
```

## 3

3.  How many rows and columns are in `yrbss_data`?

```{r}
nrow(yrbss_data)

ncol(yrbss_data)

# both rows & columns:
dim(yrbss_data)
```

## 4

4.  Which variables are numeric (quantitative)?

```{r}
glimpse(yrbss_data) 
```

- Record, bmi, and stweight are numeric, although record actually serves as a categorical variable.
- Interestingly age and grade are not.

## 5

5.  Which variables are character?

```{r}
glimpse(yrbss_data) 
```

- Age, sex, grade, race4, and race7 are character variables.

# Part 2

## 1

1.  Create a contingency table for `race4` and `race7`, with both percentages and counts. What is the difference between these two variables?

```{r}
yrbss_data |>
  tabyl(race7, race4) |>
  adorn_percentages() |>
  adorn_pct_formatting() |>
  adorn_ns() |>
  adorn_title(placement = "combined") |>
  gt()
```

`race4` groups race/ethnicity into 4 categories, while `race7` uses a finer 7-category breakdown. The table shows how each of the 7 `race7` groups maps into the 4 `race4` groups — `race4` is the coarser collapsed version of `race7`.

## 2

2.  Create a contingency table for `race4` and `race7`, with both percentages and counts, stratified by `grade`.

```{r}
yrbss_data |>
  tabyl(race7, race4, grade) |>
  adorn_percentages() |>
  adorn_pct_formatting() |>
  adorn_ns() |>
  adorn_title(placement = "combined") # gt doesn't work here
```

## 3

3.  Calculate the minimum, median, and maximum BMI and weights of participants in the YRBSS data.

There are a couple of options to do this. I chose the five number summary option.

```{r}
yrbss_data |> 
  get_summary_stats(
    bmi, stweight,
    type = c("five_number")
    ) |> 
  gt()
```

## 4

4.  Calculate the minimum, median, and maximum BMI and weights of the participants in the YRBSS data, now stratified by age.

```{r}
yrbss_data |> 
  group_by(age) |> 
  get_summary_stats(
    bmi, stweight,
    type = c("five_number")
    ) |> 
  gt()
```

# Part 3

## 1

1.  Create a histogram for the `stweight` variable. Describe the distribution shape.

```{r}
ggplot(yrbss_data, aes(x = stweight)) + 
  geom_histogram()
```

The student weights have a skewed right distribution, with a long tail for higher weights.

## 2

2.  What happens if you change `x =` to `y =` in the histogram code?

The axes are flipped and we get a horizontal histogram.

```{r}
ggplot(yrbss_data, aes(y = stweight)) + 
  geom_histogram() # axes are flipped
```

## 3

3.  Create boxplots of `stweight` stratified by sex.

The weights of males are slightly greater than for females.

```{r}
ggplot(yrbss_data, 
       aes(x = stweight, y = sex)) + 
  geom_boxplot() 
```

## 4

4.  Create a barplot of race4 categories with the y axis showing percentages.

```{r}
ggplot(data = yrbss_data, 
       aes(x = race4)) +
  # specify aesthetics within the barplot to show proportions
  geom_bar(aes(y = after_stat(prop), group = 1)) +
  # Next line converts y-axis labels to percentages instead of proportions
  scale_y_continuous(labels =  scales::percent_format())   +
  labs(y = "Percent") # default label is prop
```

## 5

5.  Visualize the missingness patterns of the `toy_data` YRBSS subset using `vis_miss()`. What missingness patterns do you see?

```{r}
toy_data <- read_excel("data/toy_data.xlsx", na = "NA")

vis_miss(toy_data)
```

We see a lot of missingness in the `text_while_driving` variable (about half missing), and some missingness for `smoked_ever`, `bullied_past_12mo`, and `race4`.

# Part 4

Perform some subsetting functions on the `yrbss_data` data:

## 1

1.  Use `filter()` to select people who are in 9th or 10th grade and who have BMI less than 25. Save the resulting data as `yrbss_filtered` tibble.

```{r}
# first let's see what the categories look like
yrbss_data |>
  tabyl(grade)

# now we filter with the "or" operator | on grade and then the "and" operator or comma on bmi
yrbss_filtered <- yrbss_data |>
  filter(grade == "9th" | grade == "10th", 
         bmi < 25
         )

# check this worked
nrow(yrbss_filtered)
yrbss_filtered |> 
  group_by(grade) |> 
  get_summary_stats(
    bmi,
    type = c("common")
    ) |> 
  gt()

# Note that output only includes 9th & 10th grade
# and max BMI is < 25
```

## 2

2.  Use `rename()` to rename 2 of the column names of `yrbss_filtered` to something more meaningful.

```{r}
names(yrbss_filtered)

yrbss_filtered2 <- yrbss_filtered |> 
  rename(ID = record, 
         weight = stweight)

names(yrbss_filtered2)
```

## 3

3.  Use `select()` to keep only the columns that denote record, sex, grade, bmi, and the 2 renamed columns.

```{r}
yrbss_filtered3 <- yrbss_filtered2 |>
  select(ID, sex, grade, bmi, weight)

# check:
glimpse(yrbss_filtered3)
```

## 4

4.  Arrange the data by bmi, starting with the highest value and ending with the least value.

```{r}
yrbss_filtered3 <- yrbss_filtered3 |>
  arrange(desc(bmi))

yrbss_filtered3
```

## 5

5.  Save the resulting data set in the file `data/practice4_data.xlsx`.

```{r}
write_xlsx(yrbss_filtered3, path = "data/practice4_data.xlsx")
```

# Part 5

*Parts 5–7 build on `yrbss_new`, the dataset we created in the slides. We rebuild it here so this solutions file runs on its own:*

```{r}
#| label: build-yrbss-new
# Rebuild yrbss_new from yrbss_data (mirrors the mutate/factor slides)
yrbss_new <- yrbss_data |>
  mutate(
    weight_lb = stweight * 2.205,   # stweight is recorded in kg
    weight_g  = stweight * 1000,
    grade     = factor(grade, levels = c("9th", "10th", "11th", "12th"))
  )
```

## 1

1.  Create a new variable `height` that calculates height based on `bmi` and `stweight`. Note the formula for BMI is weight/height\^2 (weight is in kilograms and height is in meters). Create a scatterplot of `bmi` vs `height`.

Rearranging BMI = weight/height\^2 gives height = sqrt(weight/BMI). Since `stweight` is already in kilograms, we use it directly:

```{r}
yrbss_new <- yrbss_new |>
  mutate(height = sqrt(stweight/bmi))

ggplot(yrbss_new, aes(x = height, y = bmi)) +
  geom_point()
```

The heights are reasonable for high-school students (median ~ 1.68 m), a good sanity check that we used weight in kilograms.

## 2

2.  Create a new factor variable `sex_fac` that orders the values of `sex` as Male, Female.

```{r}
yrbss_new <- yrbss_new |>
  mutate(sex_fac = factor(sex,
                          levels = c("Male", "Female")))

# check:
glimpse(yrbss_new)

yrbss_new |> tabyl(sex_fac, sex)
```

# Part 6

## 1

1.  Create a new variable called `grade_num` in `yrbss_new` that converts `grade` to numeric.

```{r}
yrbss_new |> tabyl(grade)

yrbss_new <- yrbss_new |>
  mutate(grade_num =
           case_when(
             grade == "10th" ~ 10,
             grade == "11th" ~ 11,
             grade == "12th" ~ 12,
             grade == "9th" ~ 9))

# check:
yrbss_new |> tabyl(grade, grade_num)

glimpse(yrbss_new)

```

## 2

2.  Create a new variable called `race_cat` that relabels `race4` to have shorter race category names: W (White), H/L (Hispanic/Latino), B/AA (Black or African American), Other (All other races). Make `race_cat` a factor variable with levels ordered by the size of the group in descending order.

```{r}
yrbss_new |> tabyl(race4) |> arrange(desc(n))

yrbss_new <- yrbss_new |>
  mutate(
    race_cat = case_when(
      race4 == "White" ~ "W",
      race4 == "Hispanic/Latino" ~ "H/L",
      race4 == "Black or African American" ~ "B/AA",
      race4 == "All other races" ~ "Other"
      ),
    race_cat = factor(race_cat, levels = c("W", "Other", "H/L", "B/AA"))
    )

```

::: callout-tip
Hardcoding the `levels` works here, but it breaks if the group sizes change. A more robust way to order a factor by frequency (largest group first) is `fct_infreq()` from `{forcats}` (a core tidyverse package):

```{r}
yrbss_new |>
  mutate(race_cat2 = fct_infreq(race_cat)) |>
  pull(race_cat2) |>
  levels()
```
:::

## 3

3.  Create a `tabyl` of the new variable `race_cat` and `race4` to check your work, including the ordering of the factors.

```{r}
# Check:
yrbss_new |> tabyl(race4, race_cat) |> adorn_title()
```

# Part 7

## 1

1.  Create a barplot of `race4`, faceted by `sex`.

```{r}
ggplot(yrbss_new, aes(x = race4)) +
  geom_bar() +
  facet_wrap(vars(sex)) +
  theme(axis.text.x = element_text(angle = -45, hjust = 0))
```
